---
title: "Rogue G values and other space oddities"
author: "BenjS"
date: "February 3, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", message = FALSE, warning = FALSE)
```

In this exploration of the open exoplanet database, I will:

1) Go through some basic familiarization with the data.

2) Calculate the gravitational constant in these star systems using Kepler's third law.

3) Look for structure or clustering in the data using principle component analysis (PCA).

There are a few surprises along the way--a most welcome occasion to practice our data interpretation ninja skills.

##1) Basic familiarization with the data
Data and required libraries, load yourselves!
```{r}
options(warn=-1, message=FALSE)
#Load necessary libraries
library(ggplot2)
library(FactoMineR)
library(factoextra)
library(formattable)
library(tidyr)
#Read in the data
df_raw <- read.csv("../input/oec.csv", stringsAsFactors=FALSE)
```
I'm going to replace the integer values under the star type variable ("TypeFlag") with character labels as given in the data documentation. This step is not too important but makes for a nicer graph legend farther below.
```{r}
#Replace TypeFlag integer IDs with verbal labels.
u <- df_raw$TypeFlag
df_raw$TypeFlag[which(u == 0)] <- "No known binary"
df_raw$TypeFlag[which(u == 1)] <- "P-type binary"
df_raw$TypeFlag[which(u == 2)] <- "S-type binary"
df_raw$TypeFlag[which(u == 3)] <- "No host star"
rm(u)
```
Let's familiarize ourselves with the data a bit. How many planets does this dataset cover?
```{r}
print(paste("Number of planets: ", nrow(df_raw)))
```
Quite a few. I think there were only about 2000 in the sample last time I checked several years ago. Scattered across how many star systems?
```{r}
print(paste("Number of star systems: ", length(unique(df_raw$HostStarMassSlrMass))))
```
So, 3584 planets spread across 539 star systems. How far away are these star systems?
```{r}
summary(df_raw$DistFromSunParsec)
```
From the descriptive statistics we see that these exoplanet systems are between about 1.3 and 8500 parsecs away from us (with quite a few missing values).

What sort of discovery methods are there?
```{r}
print(unique(df_raw$DiscoveryMethod))
```
Not sure what's going on with the blank ("") category, but I'll replace it with "not reported", which will make for a more aesthetically pleasing group legend in the graphs farther below. Let's see how many planets have been found by each method.
```{r}
df_raw$DiscoveryMethod[which(df_raw$DiscoveryMethod == "")] <- "not reported"
table(df_raw$DiscoveryMethod)
```
Most exoplanets have been found using the transit method, followed by the radial velocity method ("RV"). A small fraction have been found via direct imaging, microlensing, and timing. The solar system has probably been assigned NA or "" under discovery method. I'll change this to "imaging".
```{r}
u <- df_raw$ListsPlanetIsOn
solar_system <- df_raw$PlanetIdentifier[which(u == "Solar System")]
df_raw$DiscoveryMethod[which(df_raw$PlanetIdentifier %in% solar_system)] <- "imaging"
```


##2) Is the gravitational constant constant?

Let's keep the astronomers honest by checking up on Newton's gravitational constant, big G. Big G can be calculated based on Kepler's third law:
$$
\frac{T^2}{a^3} = \frac{4 \pi^2}{G(m + M)}
$$
where T is the given planet's orbital period, a is its semi-marjor axis, m is its mass, and M is its host star mass. Simply rearrange this as follows:
$$
G = \frac{4 \pi^2 a^3}{T^2 (m + M)}
$$
The exoplanet dataset includes all the variables necessary to make this calculation. Planet mass is expressed in units of Jupiter masses, while star mass is expressed in solar masses. So, first I need to do a unit conversion. Google tells me that there are 0.000954265748 Jupiter masses to a solar mass.
```{r}
conversion_factr <- 0.000954265748
df_raw$PlanetaryMassSlr <- df_raw$PlanetaryMassJpt * conversion_factr
```
Alrighty then, now that both planet and host star masses are expressed in solar units, it's time to calculate big G. I'll examine the result of this calculation be means of a density plot. If all is well in the universe then all the calculations should be equal and you'll see a tall bell-shaped curve extending over a miniscule range of values.
```{r}
df_raw$G <- df_raw$SemiMajorAxisAU^3 / df_raw$PeriodDays^2 * 4 * pi^2 /(df_raw$HostStarMassSlrMass + df_raw$PlanetaryMassSlr)

ggplot(df_raw, aes(G)) + geom_density(colour = "black", fill = "#56B4E9") + scale_x_log10(name = "Big G (log10 scale)")
```
Hmmm, we get a tall spike instead of a tall bell. Most values are reassuringly at or very near the value marked by the spike. But there are some rogue blips in the tails diverging from the constant by orders of magnitude. One of these is so extreme that I had to plot on log scale just to make the spike visible. Let's track these rogue Gs down by seeing how many observations diverge wildly from the mean (logged) value (<> 5 $\sigma$).

```{r}
df_raw$lG <- log(df_raw$G)
m_lG <- mean(df_raw$lG, na.rm = T)
s_lG <- sd(df_raw$lG, na.rm = T)
ind_rogue_hi <- which(df_raw$lG > m_lG + 5 * s_lG)
ind_rogue_lo <- which(df_raw$lG < m_lG - 5 * s_lG)
cat(paste("Number of planets with G > 5 stand. dev.: ", length(ind_rogue_hi)), paste("Number of planets with G < 5 stand. dev.: ", length(ind_rogue_lo)), sep = "\n")
```
There is one value wildly in excess of the mean, and two under it. Let's find out more about these mysterious planets.
```{r}
ind_rogue <- c(ind_rogue_hi, ind_rogue_lo)
formattable(df_raw[ind_rogue, c("PlanetIdentifier", "DiscoveryYear", "TypeFlag", "ListsPlanetIsOn", "DiscoveryMethod")])
```
The rogue exoplanets orbit single stars and were discoverd using the radial velocity method. The most recent of these discoveries, HD 39392 b, is still on the controversial list. Other than that, I don't see anything too mysterious here. Let's look at some of the physical data.
```{r}
formattable(df_raw[ind_rogue, c("PeriodDays", "SemiMajorAxisAU", "Eccentricity", "PlanetaryMassJpt", "HostStarMassSlrMass")])
```

I'm just a hack at this astronomy stuff, but the orbital period and semi-major axis of Gliese 317 c strike me as very high, and also conspicuously round. I'm guessing those are meant as upper bounds rather than an attempt at pinpoint accuracy. (Nonetheless, you'd think these scientists would report values consistent with the known laws of the universe!) The eccentricity is also very high--more like what you'd expect of a comet. As for the other two exoplanets, again, not seeing anything that raises an obvious flag.

Note that I set a pretty high outlier threshold of 5 $\sigma$. I could have pulled out more not-quite-as-extreme, but still definitely rogue, values of G had I set the threshold lower, say at 2 $\sigma$.

```{r}
ind_rogue_hi <- which(df_raw$lG > m_lG + 2 * s_lG)
ind_rogue_lo <- which(df_raw$lG < m_lG - 2 * s_lG)
cat(paste("Number of planets with G > 2 stand. dev.: ", length(ind_rogue_hi)), paste("Number of planets with G < 2 stand. dev.: ", length(ind_rogue_lo)), sep = "\n")
```
But let's move on. Hopefully someone can shed some light on this space oddity in the comments.

##3) Are we special?

Here I will look for structure or clustering in the data using PCA() from the factoMineR package.

The distance to Sun is given as NA for the solar system planets, which will result in the solar system being dropped. So, first I replace these NAs with 0.
```{r}
#Replace distance to Sun NAs with 0 for the solar system planets.
u <- df_raw$PlanetIdentifier
df_raw$DistFromSunParsec[which(u %in% solar_system)] <- 0
rm(u)
```

I perform the PCA over the physical planet and host star attributes (orbital parameters, temperature, metallicity, etc.). This excludes location data (declination, etc.). The planet age variable (AgeGry) seems to have no observations, so it is excluded. The planet surface temperature variable (SurfaceTempK) also has a lot of NAs and so is excluded.

For purposes of grouping, I will also keep the system type and discovery method variables.

I'll then drop any observations that are incomplete in these variables.
```{r}
#Keep only columns needed for this EDA
df <- df_raw[, c("PlanetIdentifier", "TypeFlag", "DiscoveryMethod", "SemiMajorAxisAU", "PeriodDays", "PlanetaryMassJpt", "RadiusJpt", "Eccentricity", "HostStarMassSlrMass", "HostStarRadiusSlrRad", "HostStarMetallicity", "HostStarAgeGyr", "HostStarTempK")]
#Drop rows with missing values
nvars <- ncol(df) - 3
ind_drop <- c()
t <- 0
for(i in 1:nrow(df)){if(sum(is.na(df[i, ])) > 0){t <- t + 1; ind_drop[t] <- i}}
df <- df[-ind_drop, ]
```
How many planets are we left with?
```{r}
print(nrow(df))
```
Wow, out of 3584 planets in the dataset, only 117 have complete data for these variables. That's still enough to do a meaningful PCA.

I'll use fviz_biplot() from the factoextra package to visualize results in the principle component space, taking advantage of the "habillage" option to group the planets by star type.
```{r}
df_pca <- df
pca_group <- as.factor(df_pca$TypeFlag)
rownames(df_pca) <- df_pca$PlanetIdentifier
df_pca <- df_pca[, c(4:ncol(df_pca))]
res <- PCA(df_pca, graph = F)
fviz_pca_biplot(res, habillage = pca_group) + ylim(-7, 5) + xlim(-8, 10)
```
For the most part, the data look pretty homogenous--basically one big cluster. However, I find it very interesting how the solar system planets diverge from the main cluster; and that this divergence is mimicked by one other system, HR 8799.

The grouping by star system type isn't too interesting. Almost all the planets belonging to binary systems have incomplete data and were thus filtered out above. Maybe this is because it is harder to measure physical parameters in binary systems.

Let's see what happens when we group by discovery method instead.
```{r}
df_pca <- df
pca_group <- as.factor(df_pca$DiscoveryMethod)
rownames(df_pca) <- df_pca$PlanetIdentifier
df_pca <- df_pca[, c(4:ncol(df_pca))]
res <- PCA(df_pca, graph = F)
fviz_pca_biplot(res, habillage = pca_group) + ylim(-7, 5) + xlim(-8, 10)
```
Huh, interesting. The solar system and HR 8799 are the only systems in the plot measured via direct imaging. And they are the only systems diverging from the main cluster.

Coincidence, or space oddity?

Note also the direction of their divergence. It's in the direction of the period and semi-major axis variable arrows, suggesting that the orbital periods and semi-major axes in these systems are considerably different from those of most discovered star systems.

Let's have a closer look at just those two variables.
```{r}
df_plot <- df[, c("PlanetIdentifier", "DiscoveryMethod", "SemiMajorAxisAU", "PeriodDays")]
df_plot <- df_plot %>% gather(Variable, Value, SemiMajorAxisAU:PeriodDays)

gg <- ggplot(df_plot, aes(Value, fill = DiscoveryMethod))
gg <- gg + geom_density(alpha = 0.4)
gg <- gg + scale_x_log10(name = "(log10 scale)")
gg <- gg + facet_wrap(~Variable, ncol = 1)
gg
```
Curiouser and curiouser: The densities are bimodal, with a high probability mode (big hump) over lower values and a relatively lower probability mode (small hump) over a range of higher values.

Kinda makes sense considering that, for transit and RV methods, planets orbiting close to their stars are much more likely to be found than planets on long orbits.

Secondly, note that there is greater agreement across discovery methods in the large modes than in the small modes. Also kinda makes sense considering that, for RV and transit methods, measurements of long orbits are presumably more difficult than of short orbits.

To get back to the motivating question though: The density of values found by direct imaging (which includes the solar system and HR 8799) is in extreme disagreement with the other methods, as suggested in the PCA biplot. Either we and HR 8799 happen to be very special systems, or there's something going on in the transit and RV (and "not reported") methods that biases those samples downwards.

I do feel very special, generally speaking, but I'm guessing it is the latter. The transit and RV methods are probably discovering disproportinately many short period exoplanets because those are the easiest to detect for those methods. Direct imaging, on the other hand, can presumably detect long period orbits with more or less the same probability that it detects short period orbits.

So, not a coincidence, but not a space oddity either.

Last question: what sort of structure do we see if we drop the imaged systems?

```{r}
ind_drop <- which(df$DiscoveryMethod == "imaging")
df_pca <- df[-ind_drop, ]
pca_group <- as.factor(df_pca$DiscoveryMethod)
rownames(df_pca) <- df_pca$PlanetIdentifier
df_pca <- df_pca[, c(4:ncol(df_pca))]
res <- PCA(df_pca, graph = F)
fviz_pca_biplot(res, habillage = pca_group) + ylim(-2.5, 8) + xlim(-6, 6)
```
It looks like some clustering does begin to emerge, with the period and semi-major axis variables again playing a key role, but this time in a way that is unrelated to the discovery method.

Enough space exploration for now. A further useful exercise might be to repeat the density plots above over the whole sample for which the period and semi-major axis data are complete. Maybe throw in eccentricity too.

You've been very kind to read this far. Comments and suggestions welcome.